S1: Extended Derivations for `A Data Driven Framework for Patient-Specific Neural Field Modelling'

### 2010-06-26 ltx/PLoS/LaTeX_Manuscript.tex
small edits to discussion and added acknowledgement

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}
With recent advances in medical imaging techniques and computing power we are entering a new age of neuroscientific research, where the amount of patient specific information is becoming incredibly detailed. In order to make the best use of this information, patient specific mathematical models should should be formed. This is particularly poignant where considering poorly understood complex systems such as large scale dynamics of the human brain. In terms of brain dynamics, there currently exists a library of neural field models that provide a theoretical explanation of various phenomena associated with various cognitive states and disease. In this paper we have created a platform for creating a patient specific neural field model. Within this platform, we propose rules for designing experiments in order to acquire the data with sufficient information and the mathematical formula to create such a model.

@@

\section*{Discussion}\label{DiscussionSection}
This paper has presented a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. As seen in Section~\ref{sec:state_and_param_results}, the results indicate unbiased estimation, where the mean parameter estimates are in good agreement with the actual values. The estimates of the connectivity kernel parameters show relatively broad distributions, which can be expected considering the system's input is modelled as a random disturbance, and the observations are corrupted by noise.

@@

To overcome the effect of noise, the use of a controlled input stimulus should be considered when applying the estimation algorithm to electrophysiological recordings. Given knowledge of the input and the appropriate model, it is expected that the variance of the parameter estimates will decrease.

Even when using an input, our results imply that we should only expect to learn distributions of parameters, rather than uncover a `true' parameter set. By designing an input signal carefully and using an evoked-potential paradigm, where stimuli are repeated many times, data could be epoched into (and averaged over) a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

@@

The current study builds on \cite{schiff2008kalman}, where the estimator is constructed to approximate a continuous field and the dimensionality of the state vector is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis methods presented in Section~\ref{SpectralAnalysisSection}. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field, with a lower dimensional state vector than would be possible otherwise.

\dean{Just need to link this to the previous bit.} The fact that the connectivity kernel is isotropic about the origin means that we are able to re-write the spatial mixing effect of the kernel on the firing rate as a convolution of the kernel on the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

\dean{Nice one Mike! Yup, we do talk a little about the inner product convolution exchange in the derivation.}

@@

\mike{again, this paragraph seems to have multiple messages. At the top is hey there are other techniques, and our is better than them. Then there's a repeat of the ``our state dimension is indie of observations'' argument which we already made above, then there's something about multi-resolution models and how two frameworks could be combined to give anisotropy. I quite like the bit about other models, but do we really need to speculate about multi-res and anisotropy? I don't think the relationship is particularly clear...}

\dean{There is definitely no repeat now. Also, I have commented out the speculative part about how to include anisotropy.}

Generating patient-specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

@@

The development of an estimation framework also has theoretical implications. The the ability to estimate parameters will allow testing of hypotheses that have been generated in theoretical studies. This has the potential to validate neural field models, and in doing so, provide new insights that may lead to a greater understanding of macro-scale neurodynamics. For example, using the proposed model-based framework we can reconstruct a neural field from data and estimate parameters to check if they correspond to theoretically derived values where certain solutions can exist. In addition, the parameters and the state-space can be explored during cognitive tasks and states of arousal such as during sleep and anaesthesia.

In this paper, an assumption was made that the chosen kernel basis functions can exactly represent the actual connectivity kernel. In future work, we plan to relax this assumption. We hope to achieve this by using the actual kernel that cannot exactly be presented by the chosen finite number of basis functions. In addition, we have assumed the neural field is isotropic and homogeneous. It is generally accepted that the intracortical connectivity structure can be accurately modelled as being isotropic. However, the assumption of homogeneity is quite strong. Future work should be focussed on developing estimation algorithms where assumption of homogeneity can be relaxed. Future work should also be targeted to include a more realistic synaptic response kernel, with finite rise and decay times, and finite, distant dependant propagation times into the parametric neural field equations.

% This paper presents a novel framework for state and parameter estimation of spatio-temporal neural fields to create patient specific models. A basis function decomposition reduces the infinite dimensional neural field (continuous space) to a finite dimensional state-space model. The frequency analysis methods presented in this paper provide a systematic procedure for specifying the sensor and basis function arrangement required to describe the significant neuro-dynamics by the reduced neural field model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch-Tung-Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. We have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic dynamics using the proposed methodology.

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

>>>>

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}
With recent advances in medical imaging techniques and computing power we are entering a new age of neuroscientific research, where the amount of patient specific information is becoming incredibly detailed. In order to make the best use of this information, patient specific mathematical models should should be formed. This is particularly poignant where considering poorly understood complex systems such as large scale dynamics of the human brain. In terms of brain dynamics, there currently exists a library of neural field models that provide a theoretical explanation of various phenomena associated with various cognitive states and disease. In this paper we have created a platform for creating a patient specific neural field model. Within this platform, we propose rules for designing experiments and derive mathematical formula to create such a model.

@@

\section*{Discussion}\label{DiscussionSection}
This paper has presented a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. As seen in Section~\ref{sec:state_and_param_results}, the results indicate accurate estimation, where the mean parameter estimates are in good agreement with the actual values. The estimates of the connectivity kernel parameters show relatively broad distributions, which can be expected considering the system's input is modelled as a random disturbance, and the observations are corrupted by noise.

@@

To overcome the effect of noise, the use of a controlled input stimulus should be considered when applying the estimation algorithm to electrophysiological recordings. Given knowledge of the input and the appropriate model, it is expected that the variance of the parameter estimates will decrease. Even when using an input, our results imply that we should only expect to learn distributions of parameters, rather than uncover a `true' parameter set. By designing an input signal carefully and using an evoked-potential paradigm, where stimuli are repeated many times, data could be epoched into (and averaged over) a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

@@

The current study builds on this previous work~\cite{schiff2008kalman}, where the estimator is constructed to approximate a continuous field and the dimensionality of the state vector is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis methods presented in Section~\ref{SpectralAnalysisSection}. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field, with a lower dimensional state vector than would be possible otherwise.

The fact that the connectivity kernel is isotropic about the origin means that we are able to re-write the spatial mixing effect of the kernel on the firing rate as a convolution of the kernel on the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

@@

Generating patient-specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interfaces or epileptic seizure prediction, where specific regions of the state-space may indicate intent of movement or imminent seizures, respectively. In addition, a patient-specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation, thereby possibly preventing seizure onset or abating seizure activity.

@@

The development of an estimation framework also has theoretical implications, where the ability to estimate parameters will allow testing of hypotheses that have been generated in computational modelling studies. This has the potential to validate neural field models, and in doing so, provide new insights that may lead to a greater understanding of macro-scale neurodynamics. For example, using the proposed model-based framework, we can reconstruct a neural field from data and estimate parameters to check if they correspond to theoretically derived values where certain solutions can exist. In addition, the parameters and the state-space can be explored during cognitive tasks and states of arousal such as during sleep and anaesthesia.

In this paper, an assumption was made that the chosen kernel basis functions can exactly represent the actual connectivity kernel. In future work, we plan to relax this assumption. We hope to achieve this by using the actual kernel that cannot exactly be presented by the chosen finite number of basis functions. In addition, we have assumed that the neural field is isotropic and homogeneous. It is generally accepted that the intracortical connectivity structure can be accurately modelled as being isotropic. However, the assumption of homogeneity is quite strong. Future work should focus on developing estimation algorithms where assumptions of homogeneity can be relaxed. Future work should also include a more realistic synaptic response kernel, with finite rise and decay times, and finite distant dependant propagation times into the parametric neural field equations.

% This paper presents a novel framework for state and parameter estimation of spatio-temporal neural fields to create patient-specific models. A basis function decomposition reduces the infinite-dimensional neural field (continuous space) to a finite-dimensional state-space model. The frequency analysis methods presented in this paper provide a systematic procedure for specifying the sensor and basis function arrangement required to describe the significant neuro-dynamics by the reduced neural field model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch-Tung-Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. We have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic dynamics using the proposed methodology.

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
This research was funded in part by the Australian Research Council (Linkage Project LP0560684). The Bionic Ear Institute acknowledges the support it receives from the Victorian State Government through the Operational Infrastructure Support Program. Dean Freestone would like to thank the University of Melbourne's Scholarships Office for the Postgraduate Overseas Research Experience Scholarship and the Harold Mitchell Foundation for a traveling scholarship and supporting this research. In addition, Dean Freestone would like to thank Dr. Mark van Rossum for hosting him whilst visiting the University of Edinburgh where this work was undertaken.
